Static contract checking for Haskell

### 2011-10-19 paper.tex
Paper: made a full comments pass through section 1.
Ouch!

Consider:
\begin{verbatim}
data List a = Nil | Cons a (List a)

@@

Then, we must state that $Nil$ never crashes (ie cannot be evaluated
to an exception) and that $Cons(x,y)$ crashes iff either $x$ or $y$
crashes. The statement $x$ crashes is encoded by the term $\cf{x}$.

@@

We also say some stuff about unreachibility but I can't think of a
good way to explain it right now.
\begin{equation*}
\forall y,ys.~Cons(y,ys) \neq \unr 
\end{equation*}
\begin{equation*}
Nil \neq \unr 
\end{equation*}

Finally, we define projections for $Cons$. It is not strictly
necessary, but it will be handy:
\begin{equation*}
\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
\end{equation*}
\begin{equation*}
\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
\end{equation*}

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.
\begin{equation*}
\forall xs.~ xs=Nil \implies notnull(xs) = false
\end{equation*}

@@

Finally, we say that one call $notnull$ with a an argument which is
not $Nil$ or $Cons$ or $\bad$ then the result is $\unr$.
\begin{align}
  \label{unrnotnull}
  \forall xs.~xs \neq \bad \land xs \neq Nil  \notag \\
  \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
  notnull(X) = \unr 
\end{align}

The translation of $head$ follows the same pattern:
\begin{align}
  \forall xs.~ x=Nil \implies head(xs) = \bad \notag\\
  \forall xs,y,ys.~ x=Cons(y,ys) \implies head(xs) = y     \label{defhead} \\
    \forall xs.~xs=\bad \to head(xs) = \bad \notag
\end{align}

\begin{align*}
 \forall xs.~xs \neq \bad \land xs \neq Nil \notag \\
 \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
 head(X) = \unr 
\end{align*}

@@

\ncnote*{Thi is  a note}{Now that we translated} everything to first-order logic, we can ask the
theorem prover if the theory formed by those formulae is
consistent, ie if $T \vdash \phi$.

Intuitively, $T$ is consistent (ie $T \not \vdash \bot$), because each
formula serves a specific purpose. Now, assume that $xs$ satisfies
$\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
because we have $\cf{xs}$ and $\lnot \cf{\bad}$. The constraint
$notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
some $y$ and $ys$. But $notnull$ is totally defined, because of
\eqref{unrnotnull}. This implies (by \eqref{defnotnull}) that there exist
$y$ and $ys$ such that $xs = Cons(y,ys)$. Recalling $\cf{xs}$, we can
now derive $\cf{y}$ and $\cf{ys}$ (by \eqref{cfcons}). But $head(xs) =
y$ because of \eqref{defhead}, and $y$ is crash-free, so we can finally
derive $\cf{head(xs)}$. QED.

@@

\begin{figure}[h]
  \centering
  \[
  \begin{array}{rclr}
    mod &:=& def_1,\dots,def_n &\\
    def &\in& \mbox{Definition} & \\
    def &:=& \texttt{data } T = K_1 \mid \cdots \mid K_n & \mbox{Data definition}\\
    &\mid& f \in c & \mbox{Contract claim}\\
    &\mid& f~\vec{x}~=~e & \\
    &\raisebox{ 0.18 cm }{$\mid$}& \multicolumn{2}{c}{
      \begin{array}{rcl}
        \hspace{-0.2cm} f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
         && K_1(\vec{x}_1) \to e_1 \mid \cdots \mid K_n(\vec{x}_n) \to e_n \\
       \end{array} 
       } \\
  \end{array} \]

@@

\subsection{Crash-freeness}
Note that $\cfc$ represents two things: it can be a contract, as in $f
\in \cfc$ or a special formula in first-order logic $\cf{f}$.

@@

\subsection{Contract checking -- Recursive case}
If the function $f=e$ is recursive, then we ask the theorem prover the
following:

>>>>

Consider:\note[nc]{Unlike the POPL '09 paper, $CF(e)$ is not implied by $e \in \{x|p\}$.}
\begin{verbatim}
data List a = Nil | Cons a (List a)

@@

Then, we must state that $Nil$ never crashes (ie cannot be evaluated
to an exception) and that $Cons(x,y)$ \add[nc]{never} crashes iff \change[nc]{either}{both} $x$ \change[nc]{or}{and} $y$
\add[nc]{never} crash. The statement $x$ \add[nc]{never} crashes is encoded by the term $\cf{x}$\note[nc]{Made more precise in section \ref{sec:cf}}.

@@

We also say some stuff about \annote[nc]{unreachibility}{??? e.g. \[
  CF(fst(x,BAD)) \iff CF(x) \] as in section \ref{sec:cf}?} but I
can't think of a good way to explain it right now.
\begin{equation*}
\forall y,ys.~Cons(y,ys) \neq \unr 
\end{equation*}
\begin{equation*}
Nil \neq \unr 
\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\Phi_4$) say that constructors are ``lazy'' or ``total''.}

Finally, we define projections for $Cons$. \annote[nc]{It is not
  strictly necessary, but it will be handy:}{??? Why is it handy? It
  doesn't seem to play any role in formal part of this paper (besides
  being specified in axioms $\Phi_1$ (projections)). In the current
  implementation it's used to avoid introducing existentially
  quantified variables in the $\bigvee_i$ part of $\dtrans{}$ for
  functions defined by case expressions. Equation \eqref{unrnotnull}
  below uses it that way. ??? Does Equinox care?}
\begin{equation*}
\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
\end{equation*}
\begin{equation*}
\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
\end{equation*}\note[nc]{??? Why not the simpler \[\forall y,ys.~sel_{2,Cons}(Cons(y,ys)) = ys\]? Same question in Figure \ref{dtrans}.  (Equinox does have a built in notion of equality, e.g. running equinox on
\center{\texttt{fof(\_,theorem,? [X,Y] : (X = Y \& f(X) != f(Y))).}}\\
 results in ``Unsatisfiable'', so support for equality reasoning is not the reason.)}
%% \[
%% \text{\texttt{\$ echo "fof(\_,theorem,? [X,Y] : (X = Y & f(X) != f(Y)))."}} \\
%%                  \text{\texttt{| equinox /dev/stdin}}}}
%% \]

Now we translate the $null$ function. Note that the symbols $true$ and
$false$ are the representation of the data constructors $True$ and
$False$ in Haskell, not the boolean values $\top$ and $\bot$ in our
logic.\note[nc]{XXX, TODO: use $True$ and $False$ in the
  text.  The lowercase $true$ and $false$ is an artifact of the implementation, i.e. Equinox accepts TPTP format, which doesn't allow bare uppercase function symbols.  Moreover, the lower case here is inconsistent with the rest of the presentation, e.g. $Cons$ and $Nil$ above.}
\begin{equation*}
\forall xs.~ xs=Nil \implies notnull(xs) = false
\end{equation*}\note[nc]{Here we might ask the same question as above: why not use the simpler definition
\[
\forall notnull(Nil) = false
\]? The answer is that this is an expansion of the $etrans{e} = K_i(\vec{x_i}) \to a = \etrans{e_i}$ part of the $\dtrans{}$ rule for functions defined by cases.  I.e., the scrutinee $Nil$ is, in general, an expression involving, but not necessarily equal to, the arguments to the function.  In this special case we could remove the implication, but not in general.  The $a = \etrans{e_i}$ part is missing above, but I guess that's because this text was not updated for the $\min$-translation.}

@@

Finally, we say that one call $notnull$ with a an argument which is
not $Nil$ or $Cons$ or $\bad$ then the result is $\unr$.
\begin{align}
  \label{unrnotnull}
  \forall xs.~xs \neq \bad \land xs \neq Nil  \notag \\
  \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
  notnull(xs) = \unr 
\end{align}

The translation of $head$ follows the same pattern:
\begin{align}
  \forall xs.~ x=Nil \implies head(xs) = \bad \notag\\
  \forall xs,y,ys.~ xs=Cons(y,ys) \implies head(xs) = y     \label{defhead} \\
    \forall xs.~xs=\bad \to head(xs) = \bad \notag
\end{align}\note[nc]{??? Why not use the same project+construct trick from the last equation?  I.e., why not
\[
\forall xs.~ xs=Cons(sel_{Cons,1}(xs),sel_{Cons,2}(xs)) \implies head(xs) = sel_{Cons,1}(xs)
\]?  Because it's ugly?  Would Equinox care?}

\begin{align*}
 \forall xs.~xs \neq \bad \land xs \neq Nil \notag \\
 \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
 head(xs) = \unr 
\end{align*}

@@

Now that we translated everything to first-order logic, we can ask the
theorem prover if the theory formed by those formulae \change[nc]{is
consistent}{entails the contract}, ie if $T \vdash \phi$.

Intuitively, $T$ is consistent (ie $T \not \vdash \bot$), because each
formula serves a specific purpose.\note[nc]{

  A more precise consistency argument by exhibiting a model for $T$:
  in the model we define $\min{}$ to be the trivial predicate,
  i.e. $\forall x. \min{x}$.  So, we can drop all the $\min{}$'s.  The
  model for the resulting formulas is a language with the syntax used
  in this paper ($\mathcal{H}'$ below).

  The carrier consists of all closed (i.e. variable free / grounded)
  programs that can be written using the functions and constructors in
  $T$, using the expression syntax of $\mathcal{H}'$ below, and
  evaluated as described in the following paragraphs.  We assume $T$
  comes from a ``well-typed'' source program $P$, i.e. s.t. the
  Haskell program corresponding to $P$ is well-typed.

  The semantics are what you would expect for ``well-typed''
  terminating expressions (again, ``well-typed'' in the sense that the
  corresponding Haskell expression is well-typed in an environment
  including $P$'s decls). For example, \texttt{Cons(Z,Nil)} is
  ``well-typed'' but \texttt{Cons(Nil,Z)} is ``ill-typed'' in a the
  context of a program declaring List and Nat types.

  In general, scrutinizing a \unr{} yields \unr, as does scrutinizing
  a terminating expression with the wrong constructor head
  (e.g. \texttt{tail(S Z) =} \unr{}). Scrutinizing an ill-typed
  expression with a constructor head of the right type evaluates as
  specified by the pattern (e.g. \texttt{tail(Cons(Nil,Z)) = Z}),
  unless there is no corresponding pattern (``non-exhaustive
  pattern''), in which case the result is \bad{}.  NB: wrong-type and
  no-corresponding-pattern are treated very differently.  The
  intuition is that the former never happens in a well-typed program.

  Applying \unr{} or \bad{} as a function results in \unr{} or \bad{},
  respectively.

  Term constructors have disjoint ranges, and are total.

  I think that covers functions and datatypes, but it remains to argue
  that the axioms for contracts are consistent \ldots.

  \ldots but maybe this is tricky?  Modeling $CF(e)$ seems easy,
  i.e. just ask the oracle used above for evaluation if $e$ is \bad{}
  or not.  But recursion seems problematic, since we get to assume in
  $T$ that the contract holds in the recursive position (Section
  \ref{sec:recursiveContracts}), and so it seems we need to argue that
  \emph{all} expressable contracts are consistent.

  Here's a flawed contradiction ``proof'': consider
  \[
  e = e ::: CF \&\& \neg CF.
  \]
  Now, $T$ will contain $CF(ep) \land \neg CF(ep)$, and were hosed.
  This is flawed because there is no $\neg$ in the annotated
  contracts.

  OK, and actually this might be easy: we simply interpret all
  recursive function symbols (the $fp$'s for recursive $f$'s) as
  \unr{}, and use the fact that forall $x$, $\unr{}~x = \unr$.  Now,
  an easy induction on contract translations (Figure \ref{ctrans})
  gives \[ \emptyset \vdash \ctrans{\unr \in c}
  \]
  for all contracts $c$, and we're done!  Well, this is really just
  the base case, where all contracts in $T$ are for otherwise
  undefined recursion symbols ($fp$'s), but the inductive case is just
  as easy: given a consistent $T$, which may include contracts for
  defined symbols, we can always extend it to a consistent $T'$ by
  adjoining constracts for undefined symbols.

}

Now, assume that $xs$ satisfies
$\cf{xs}$ and $notnull(xs) = true$.  We can derive that $xs \neq \bad$
because we have $\cf{xs}$ and \annote[nc]{$\lnot \cf{\bad}$}{Obviously true, given the definition of $CF$, but it's worth noting that this plays a key part in the axiomitization of $CF$, which is ommited in this draft (so far ...)}. The constraint
$notnull(xs) = true$ doesn't directly imply that $xs = Cons(y,ys)$ for
some $y$ and $ys$. But $notnull$ is totally defined, because of
\eqref{unrnotnull}. This implies (by \eqref{defnotnull}) that there exist
$y$ and $ys$ such that $xs = Cons(y,ys)$. Recalling $\cf{xs}$, we can
now derive $\cf{y}$ and $\cf{ys}$ (by \eqref{cfcons}). But $head(xs) =
y$ because of \eqref{defhead}, and $y$ is crash-free, so we can finally
derive $\cf{head(xs)}$. QED.

@@

\begin{figure}[h]
  \centering
  \[
  \begin{array}{rclr}
    mod &:=& def_1,\dots,def_n &\\
    def &\in& \mbox{Definition} & \\
    def &:=& \texttt{data } T = K_1 a_1 \mid \cdots \mid K_n a_n& \mbox{Data definition}\\
    &\mid& f \in c & \mbox{Contract claim}\\
    &\mid& f~\vec{x}~=~e & \\
    &\raisebox{ 0.18 cm }{$\mid$}& \multicolumn{2}{c}{
      \begin{array}{rcl}
        \hspace{-0.2cm} f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
         && K_1(\vec{x}_1) \to e_1 \mid \cdots \mid K_n(\vec{x}_n) \to e_n \\
       \end{array} 
       } \\
  \end{array} \]

@@

\subsection{Crash-freeness}\label{sec:cf}
Note that $\cfc$ represents two things: it can be a contract, as in $f
\in \cfc$ or a special formula in first-order logic $\cf{f}$.

@@

\subsection{Contract checking -- Recursive case}\label{sec:recursiveContracts}
If the function $f=e$ is recursive, then we ask the theorem prover the
following:
